Determinism, Complexity, and Predictability in Computer Performance

### 2014-04-12 results.tex
fixed up results.tex

%BRAINSTORMING 
%\begin{itemize}
%\cmark\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]] 
%\cmark\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this. [[talk about again here but justify in information theory section]]
%\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.

%\cmark\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured. And a constant, linear and nonlinear prediction strategy all fail. We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]

%\cmark\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]

%\cmark\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method. That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy. This gives a practitioner leverage on when to give up and when to keep working. [[use in this section as bridge to next section]]

@@

%\subitem [[Simply reminders not to be included]]WPE can quantify when
%a noisy real-valued time series is predictable.[[I am unsatisfied
%with predictable in this sentence, need a better word to say "better
%than random walk" or ``able to be forecast effectively" or ``has the
%structural capacity to transmit information in a way that the time
%series can be effectively forecast"

@@

%\subitem [[Simply reminders not to be included]]We will have shown
%that the existence whether linear or nonlinear is picked up on with
%WPE but this point gets at whether the prediction model can use the
%structure or not (linear can't use nonlinear structure). The right to
%left shifts in \col and some of the \svd regimes and the lack of
%shift in \gcc illustrate this nicely.

%\subitem [[Simply reminders not to be included]]This is the linear vs
%nonliear vs random we see with the right to left shifts with lower
%complexity time series.

@@

%\item Quantifying structured and unstructured complexity is
%nontrivial in the case of real-valued noisy time series but WPE does
%this. [[talk about again here but justify in information theory
%section]]

@@

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. $MASE$ plots for all 360
experiments.
\begin{figure}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=\columnwidth]{figs/predictions_vs_entropy}
\caption{WPE vs. $MASE$ for all trials, methods, and systems. \svdone,
  \svdthree, and \svdfive are omitted from the top plot for scale
  reasons. {\color{red} All three plots need to have the same
    horizontal scale!}
% 
% ; their $MASE$ scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively. 
% 
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The dashed line is the
same as in the previous figure.
% 
%  [[All traces except those from \svd$_1$ lie above the line,
%      indicating that LMA is better suited prediction method for the
%      computer traces considered.]]  [[This comes {\bf long} before the
%      associated discussion in the text, and it's not what this figure
%      is trying to show here.]]
% 
}
    \label{fig:wpe_vs_mase_all}
\end{figure} 
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding system in any of the three plots in
Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
much across trials.  For most traces, the variance in $MASE$ scores is
low as well, resulting in small, tight clusters.  In some
cases---ARIMA predictions of {\tt col\_major}, for instance---the
$MASE$ variance is larger, which spreads out the clusters
horizontally.  The distribution of the cluster {\sl positions} is
quite different in the three plots.  The mean $MASE$ scores of
predictions generated with the nonlinear LMA method are generally
closer to the dashed line; the ARIMA method clusters are more
widely spread, the \naive ~clusters even more so.  This geometric
validation of the second conjecture in this paper is discussed in more
details in the following paragraphs.

@@

The dashed line in the Figures is a heuristic that captures the
argument at the end of the previous paragraph.  It can allow
practitioners to recognize when a prediction method is not well
matched to the task at hand: that is, when the time series has more
predictive structure than the method is able to use.  When the
predictions are well below that line (as in the top and middle images
in Figure~\ref{fig:wpe_vs_mase_all}), one should try a different
method.

Note that the \col points (blue {\color{blue}$+$} icons) are clustered
very tightly in the lower left quadrant of the LMA and \naive ~$MASE$
vs. WPE plots, but spread out horizontally in the ARIMA plot.
{\color{red}[[We need to come up with a hypothesis about this spread,
      even if it's totally tentative.  Was it that the random-walk got
      really good in a few of the \col trials?  Or that ARIMA got
      really bad in those trials?  In either case, why did that
      happen---what was it about those trials that made ARIMA bark
      and/or random walk do well?]]} The position of the LMA
cluster---well to the left of the other two, near the dashed
line---reflects this method's ability to capture and exploit the
structure that is present in this signal.

@@

\svdone behaves very differently than the other seven systems in this
study.  Though its weighted permutation entropy is very high ($0.957
\pm 0.016$), two of the three prediction methods do quite well,
yielding $MASE$ scores of $0.827 \pm 0.076$ (LMA) and $0.714 \pm
0.075$ (ARIMA).  This pushes the corresponding clusters of points in
the bottom two images in Figure~\ref{fig:wpe_vs_mase_all} well above
the trend followed by the other seven clusters.  The $MASE$ scores of
the predictions of this system produced by the \naive ~method,
however, are highly inconsistent: $2.676 \pm 4.328$.  The majority of
the blue diamond-shaped points on the top plot in the figure are
clustered near a $MASE$ score of 0.6, but three of the 15 \svdone
trials posed challenges to this prediction method, resulting in $MASE$
scores $\approx 2.5$.

These effects, we believe, are artifacts of the way $MASE$ is
calculated, as well as the nonstationarity properties of the \svdone
signal.  Recall that $MASE$ scores are scaled \emph{relative to a
  random-walk forecast}, which uses the current value as the
prediction.  This strategy works very badly on signals with frequent,
large, rapid transitions.  Consider a signal that oscillates from one
end of its range to the other at every step.  A signal like this will
have a low WPE, much like {\tt col\_major}.  However, a random-walk
forecast of this signal will be 180 degrees out of phase with the true
continuation.  Since random-walk error appears in the denominator of
the $MASE$ score, this effect can shift points leftwards on a WPE
vs. $MASE$ plot, and that is exactly why the \svdone clusters for
ARIMA and LMA are above the dashed line in
Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
closeup in Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure} 
is not quite to the level of the worst-case signal described above,
but it still poses a serious challenge to random-walk prediction.  It
is dominated by a white-noise regime (between $\approx$1.86 and
$\approx$1.88 on the vertical scale in the Figure), punctuated by
short excursions above 1.9.  In the former regime, which makes up more
than 80\% of the signal, there are frequent dips to 1.82 and
occasional larger dips below 1.8.  These single-point dips are the
bane of random-walk forecasting.  In this particular case, roughly
40\% of the forecasted points are off by the width of the associated
dip, which skews the associated $MASE$ scores.  Signals like this are
also problematic for the \naive ~prediction strategy, since the
outliers have significant influence on the mean.  This compounds the
effect of the skew in the scaling factor and creates a large spread in
the \svdone $MASE$ values.  {\color{red} We need to think about the
  three outliers and put in a sentence here about them.}  For all of
these reasons, we view \svdone as an outlier and exclude it from the
fit calculation of the dashed line in
Figure~\ref{fig:wpe_vs_mase_best}.

>>>>

%BRAINSTORMING 
%\begin{itemize}
%\cmark\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.
%\cmark\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this. 
%\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.

%\cmark\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured. And a constant, linear and nonlinear prediction strategy all fail. We should be able to conclude that guessing random values is the best we can do as is shown by MASE

%\cmark\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction

%\cmark\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method. That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy. This gives a practitioner leverage on when to give up and when to keep working.

@@

%\subitem Simply reminders not to be included]]WPE can quantify when
%a noisy real-valued time series is predictable.I am unsatisfied
%with predictable in this sentence, need a better word to say "better
%than random walk" or ``able to be forecast effectively" or ``has the
%structural capacity to transmit information in a way that the time
%series can be effectively forecast"

@@

%\subitem Simply reminders not to be included]]We will have shown
%that the existence whether linear or nonlinear is picked up on with
%WPE but this point gets at whether the prediction model can use the
%structure or not (linear can't use nonlinear structure). The right to
%left shifts in \col and some of the \svd regimes and the lack of
%shift in \gcc illustrate this nicely.

%\subitem Simply reminders not to be included]]This is the linear vs
%nonliear vs random we see with the right to left shifts with lower
%complexity time series.

@@

%\item Quantifying structured and unstructured complexity is
%nontrivial in the case of real-valued noisy time series but WPE does
%this. talk about again here but justify in information theory
%section]]

@@

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. $MASE$ plots for all 360
experiments.
\begin{figure}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=\columnwidth]{figs/predictions_vs_entropy3}
\caption{WPE vs. $MASE$ for all trials, methods, and systems. \svdone,
  \svdthree, and \svdfive are omitted from the top plot for scale
  reasons.
% 
% ; their $MASE$ scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively. 
% 
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The dashed line is the
same as in the previous figure.
}
    \label{fig:wpe_vs_mase_all}
\end{figure} 
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding system in any of the three plots in
Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
much across trials.  For most traces, the variance in $MASE$ scores is
low as well, resulting in small, tight clusters.  In some
cases---ARIMA predictions of {\tt col\_major}, for instance---the
$MASE$ variance is larger, which spreads out the clusters
horizontally.  The distribution of the cluster {\sl positions} is
quite different in the three plots.  The mean $MASE$ scores of
predictions generated with the nonlinear LMA method are generally
closer to the dashed line; the ARIMA method clusters are more
widely spread, the \naive ~clusters even more so.  This geometric
validation of the second conjecture in this paper is discussed in more
details in the following paragraphs.

@@

The dashed line in the Figures is a heuristic that captures the
argument at the end of the previous paragraph.  It can allow
practitioners to recognize when a prediction method is not well
matched to the task at hand: that is, when the time series has more
predictive structure than the method is able to use.  When the
predictions are well below that line (as in the top and middle images
in Figure~\ref{fig:wpe_vs_mase_all}), one should try a different
method.  Conversely, the position of the LMA cluster---well to the
left of the other two, near the dashed line---reflects this method's
ability to capture and exploit the structure that is present in this
signal.

Note that the \col points (blue {\color{blue}$+$} icons) are clustered
very tightly in the lower left quadrant of the LMA and \naive ~$MASE$
vs. WPE plots, but spread out horizontally in the ARIMA plot.  This is
because of the way that these models are built.  If a KPSS test of the
time series in question indicates that it is nonstationary, an
integration term is added to the model.  According to this test, the
\col time series is right on the edge of nonstationarity, so the ARIMA
procedure sometimes omits the integration term, which increases the
error, thereby spreading out the points.

@@

\svdone behaves very differently than the other seven systems in this
study.  Though its weighted permutation entropy is very high ($0.957
\pm 0.016$), two of the three prediction methods do quite well on this
signal, yielding $MASE$ scores of $0.827 \pm 0.076$ (LMA) and $0.714
\pm 0.075$ (ARIMA).  This pushes the corresponding clusters of points
in the bottom two images in Figure~\ref{fig:wpe_vs_mase_all} well
above the trend followed by the other seven signals.  The $MASE$
scores of the predictions that were produced by the \naive ~method for
this system, however, are highly inconsistent.  The majority of the
blue diamond-shaped points on the top plot in
Figure~\ref{fig:wpe_vs_mase_all} are clustered near a $MASE$ score of
0.6.  In five of the 15 \svdone trials, however, there were step
changes in the signal.  The \naive ~method has a very difficult time
with signals like this, particularly if there are multiple step
changes.  This raised the $MASE$ scores of these trials, pushing the
corresponding points to the right, and in turn raising both the mean
and variance of this set of trials.

The effects described in the previous paragraph are exacerbated by the
way $MASE$ is calculated.  Recall that $MASE$ scores are scaled
\emph{relative to a random-walk forecast}, which uses the current
value as the prediction.  This strategy works very badly on signals
with frequent, large, rapid transitions.  Consider a signal that
oscillates from one end of its range to the other at every step.  A
signal like this will have a low WPE, much like {\tt col\_major}.
However, a random-walk forecast of this signal will be 180 degrees out
of phase with the true continuation.  Since random-walk error appears
in the denominator of the $MASE$ score, this effect can shift points
leftwards on a WPE vs. $MASE$ plot, and that is exactly why the
\svdone clusters for ARIMA and LMA are above the dashed line in
Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
closeup in Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure} 
is not quite to the level of the worst-case signal described above,
but it still poses a serious challenge to random-walk prediction.  It
is dominated by a noisy regime (between $\approx$1.86 and
$\approx$1.88 on the vertical scale in the Figure), punctuated by
short excursions above 1.9.  In the former regime, which makes up more
than 80\% of the signal, there are frequent dips to 1.82 and
occasional larger dips below 1.8.  These single-point dips are the
bane of random-walk forecasting.  In this particular case, roughly
40\% of the forecasted points are off by the width of the associated
dip, which skews the associated $MASE$ scores.  Signals like this are
also problematic for the \naive ~prediction strategy, since the
outliers have significant influence on the mean.  This compounds the
effect of the skew in the scaling factor and exacerbates the spread in
the \svdone $MASE$ values.  For all of these reasons, we view \svdone
as an outlier and exclude it from the fit calculation of the dashed
line in Figure~\ref{fig:wpe_vs_mase_best}.
